---
title: "ABS Strike Zone Comparison EDA"
author: "Charlie Braverman, Beili Chou, Andrew Hack"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction Charlie Test

Major League Baseball continues embracing technology by implementing an automated ball-strike (ABS) challenge system in spring training this year. While it will not be used in 2025, this system could be implemented in real games as soon as next year.. With a robotic strike zone, the league has changed its definition of the strike zone’s height to be 27% to 53.5% of a batter’s height. This marks a departure from the old rule from the midpoint between the shoulders and the top of the uniform pants to the top of the knees. While the 2025 season’s strike zone may be less impacted because human umpires will still be responsible for calling balls and strikes, the new strike zone may impact different players to different degrees. A key question for players, managers, and MLB front offices arises: which players stand to benefit from this change, and who might suffer?

The most obvious factor to investigate first is height. A taller player could see their effective strike zone shift more dramatically than a shorter counterpart. However, other elements, such as batting stance or body proportions, may also influence hitter and pitcher outcomes. By analyzing historical pitch-tracking data from Statcast, we will try to predict how the new strike zone rules will affect both hitters and pitchers.

## Conventions

with_underscores

Variables: nouns

Functions: verbs

Automated ball-strike challenge system -\> ABS

Major League Baseball -\> MLB

## Packages & Functions

Uncommon Packages: baseballr @baseballr

baseballr is a package written for R focused on baseball analysis. It
includes functions for scraping various data from websites, such as
FanGraphs.com, Baseball-Reference.com, and baseballsavant.mlb.com. It
also includes functions for calculating metrics, such as wOBA, FIP, and
team-level consistency over custom time frames.

You can read more about some of the functions and how to use them at its
official site, <http://billpetti.github.io/baseballr/index.html>.

Sourced Functions: scrape_statcast() https://jacobrichey.github.io/2020-06-06-Build-a-Statcast-Database-in-R/

The scrape_statcast() function is a custom function that scrapes pitch-by-pitch data from Statcast for a given season. It takes a season as an argument and returns a data frame with all the pitch-by-pitch data for that season. The function uses the baseballr package to scrape the data from baseballsavant.mlb.com. This is necessary since baseballr scraping functions only return up to 25000 rows of data, and we need to scrape all the data for a season.

```{r load pkgs, warning=FALSE, message=FALSE}
library(tidyverse)
library(baseballr)
```


```{r load functions, warning=FALSE, message=FALSE}
# scrape data from Statcast and write to csv
scrape_statcast <- function(season) {
  
  # create weeks of dates for season from mar - nov
  # includes spring training + postseason
  dates <- seq.Date(as.Date(paste0(season, '-03-01')),
                    as.Date(paste0(season, '-12-01')), by = 'week')
  
  date_grid <- tibble(start_date = dates, 
                      end_date = dates + 6)
  
  # create 'safe' version of scrape_statcast_savant in case week doesn't process
  safe_savant <- safely(scrape_statcast_savant)
  
  # loop over each row of date_grid, and collect each week in a df
  payload <- map(.x = seq_along(date_grid$start_date), 
                 ~{message(paste0('\nScraping week of ', date_grid$start_date[.x], '...\n'))
                   
                   payload <- safe_savant(start_date = date_grid$start_date[.x], 
                                          end_date = date_grid$end_date[.x], type = 'pitcher')
                   
                   return(payload)
                 })
  
  payload_df <- map(payload, 'result')
  
  # eliminate results with an empty dataframe
  number_rows <- map_df(.x = seq_along(payload_df), 
                        ~{number_rows <- tibble(week = .x, 
                                                number_rows = length(payload_df[[.x]]$game_date))}) %>%
    filter(number_rows > 0) %>%
    pull(week)
  
  payload_df_reduced <- payload_df[number_rows]
  
  combined <- payload_df_reduced %>%
    bind_rows()
  
  return(combined)
}
```
## Data

Statcast provides pitch-by-pitch data for every game played in MLB. Our
approach will be to pull pitch data and join with another data set that
holds: batter_id, batter_name, and batter_height. This full data set
will allow for us to sort by batter produce the new rulebook strike zone
for each. 

Further documentation on variable names and their meanings can be found here: https://baseballsavant.mlb.com/csv-docs

Biofile data is obtained from https://www.retrosheet.org/biofile.htm, specifically the biofile0 since it contains heights in inches.

The output data frame from this chunk will include all 113 variables from the Statcast data, as well as the batter's height in inches, strike zone top/bottom converted to inches, and the new ABS zone top/bottom in inches for each batter based on their height.

One assumption that must be made is that batting stance will not affect the strike zone. This is a simplification that will be made for the sake of this analysis. In the future, maybe Statcast data will be used to record batting stance height, and this assumption can be revisited.

```{r load data, results='hide', collapse=TRUE, echo=FALSE, warning = FALSE, message = FALSE}
# tracking time for optimizing code efficiency
# If scraping a season, this chunk should take about 25-30 minutes to run (oof).
t <- Sys.time()

# scrape data from Statcast for 2024 season (Only run if data is not present in data file, or if you need to download another season.)
full_pitch_df <- scrape_statcast(season = 2024)
write.csv(full_pitch_df, "data/full_pitch_df.csv")

# scrape data from Statcast on batters to build height data
batter_mlbamid_df <- scrape_statcast_savant_batter_all(start_date = "2024-01-01", end_date = "2024-09-29") %>% 
  select(batter, player_name)

# loop that adds key_retro to batter_mlbamid_df by using function playername_lookup() on each row, inputting that row's "batter" column as the argument, and selecting only key_retro to add to the data set
batter_mlbamid_df$key_retro <- NA

player_id_key_df <- chadwick_player_lu()

player_id_key_df <- player_id_key_df %>% select(key_mlbam, key_retro) %>% filter(!is.na(key_mlbam))

for (i in 1:nrow(batter_mlbamid_df)) {
  batter_mlbamid_df$key_retro[i] <- player_id_key_df %>% filter(key_mlbam == batter_mlbamid_df$batter[i]) %>% select(key_retro)
}

# turn list of key_retro into character
batter_mlbamid_df$key_retro <- as.character(batter_mlbamid_df$key_retro)

# read in biofile data and join with batter_mlbamid_df to get batter height
biofile_df <- read_csv("data/biofile_df.csv") %>% select(id, height)

batter_height_df <- batter_mlbamid_df %>% full_join(biofile_df, by = c("key_retro" = "id")) %>% select(batter, player_name, height) %>% distinct()

# join full_pitch_df with batter_height_df to get batter height in pitch data
full_pitch_df <- full_pitch_df %>% left_join(batter_height_df, by = c("batter" = "batter")) %>% rename(batter_height = height, batter_name = player_name.y, pitcher_name = player_name.x)

# changing sz_bot and top and plate_x/z to inches
full_pitch_df$sz_bot <- full_pitch_df$sz_bot * 12
full_pitch_df$sz_top <- full_pitch_df$sz_top * 12

full_pitch_df$plate_x <- full_pitch_df$plate_x * 12
full_pitch_df$plate_z <- full_pitch_df$plate_z * 12

# calculate new strike zone top and bottom based on batter height
full_pitch_df <- full_pitch_df %>% mutate(abs_sz_bot = .27 * batter_height, abs_sz_top = .535 * batter_height)

# calculate sz height and abs sz height
full_pitch_df <- full_pitch_df %>% mutate(sz_height = sz_top - sz_bot, abs_sz_height = abs_sz_top - abs_sz_bot)

# end time tracking
Sys.time()-t
```

Taking a look at pitch location data, which is broken down into x (horizontal) and z (vertical) coordinates:

```{r pitch loc descriptives}

# density plot showing plate_x and plate_z, with the strike zone overlaid, and alpha to show density, with the rectangle just an outline
# Calculate average sz_top and sz_bot
avg_top <- mean(full_pitch_df$sz_top, na.rm = TRUE)
avg_bot <- mean(full_pitch_df$sz_bot, na.rm = TRUE)

pitch_loc_plot <- ggplot(full_pitch_df, aes(x = plate_x, y = plate_z)) +
  # Use stat_density2d to compute 2D density; fill is mapped to ..level..
  stat_density2d(
    aes(fill = after_stat(level)),
    geom = "polygon",
    alpha = 0.5
  ) +
  # Rename the legend label for fill
  scale_fill_continuous(name = "Density") +  geom_rect(
    xmin = -8.5, xmax = 8.5,
    ymin = avg_bot, ymax = avg_top,
    fill = NA, color = "red",
    inherit.aes = FALSE
  ) +
  coord_fixed() +
  labs(title = "Pitch Location Density", x = "Horizontal location (inches)", y = "Vertical location (inches)")


pitch_loc_plot

# summary statistics for plate_x and plate_z
summary(full_pitch_df$plate_x)

summary(full_pitch_df$plate_z)
```

And taking a look at how much the strike zone height varies across different batters:

```{r strike zone descriptive}
# plot with overlay of average strikezone for every batter height
sz_height_plot <- ggplot(full_pitch_df, aes(x = batter_height, y = sz_height)) +
  geom_point() + geom_smooth(method = "gam") +
  labs(title = "Height vs. Strikezone Height", x = "Batter height (inches)", y = "Strikezone height (inches)")

sz_height_plot
# summary statistics for sz_top and sz_bot
summary(full_pitch_df$sz_top)
summary(full_pitch_df$sz_bot)
```

## EDA

Some initial thoughts:

How does player height currently relate to strike zone bot/top?

#### Exploring strikezone sizes:

```{r new strikezone}
# group by batter height and calculate average strike zone top and bottom
full_pitch_df %>% group_by(batter_height) %>% summarise(avg_abs_sz_top = mean(abs_sz_top), avg_abs_sz_bot = mean(abs_sz_bot), avg_sz_top = mean(sz_top), avg_sz_bot = mean(sz_bot), avg_sz_height_diff = mean(sz_height)-mean(abs_sz_height))

# graph of sz height vs batter_height, with the new strike zone height as a reference line
sz_height_vs_height <- ggplot(full_pitch_df, aes(x = batter_height, y = sz_height)) +
  geom_point() + geom_smooth(method = "gam") + geom_line(aes(x = batter_height, y = abs_sz_height), linetype = "dashed") +
  labs(title = "Height vs. Strikezone Height", x = "Batter height (inches)", y = "Strikezone height (inches)")

sz_height_vs_height
```
The new ABS zone is proportional to batter height, so it acts linearly. One takeaway is that batting stance can explain the deviation in strike zone height for each batter. However, an interesting note is that the ABS strike zone height is significantly smaller than the average current strike zone height at each batter height. Batting stance should theoretically cause current strike zone height to be smaller than ABS strike zone height, but this is not the case. Is the codification of the strike zone going to be a massive change for batters?

Not all batters and pitchers will be affected equally by the new strike zone. Some batters may benefit from the new strike zone, while others may suffer. We can calculate the difference in strike zone area between the current and ABS strike zones to see how much each batter's strike zone will change. We can also calculate the batting average on balls in zone (BABIZ) for each batter in the current and ABS strike zones to see how their performance might change. Obviously, players may change their approach to hitting based on the new strike zone, but this will give us a starting point to see who might benefit or suffer the most. 

#### Players of Interest

Some players are notorious for their height, such as Aaron Judge and Jose Altuve. We can look at how their strike zones will change and how their performance might be affected.

```{r player of interest}
# calculate strike zone area for current and ABS strike zones for all players
full_pitch_df <- full_pitch_df %>% mutate(sz_area = sz_height * 17, abs_sz_area = abs_sz_height * 17)

# create global variables for average abs strike zone top and bottom for Judge and Altuve
judge_avg_abs_top <- full_pitch_df %>% filter(batter_name=="Judge, Aaron") %>% summarise(avg_abs_top = mean(abs_sz_top))
judge_avg_abs_bot <- full_pitch_df %>% filter(batter_name=="Judge, Aaron") %>% summarise(avg_abs_bot = mean(abs_sz_bot))

altuve_avg_abs_top <- full_pitch_df %>% filter(batter_name=="Altuve, Jose") %>% summarise(avg_abs_top = mean(abs_sz_top))
altuve_avg_abs_bot <- full_pitch_df %>% filter(batter_name=="Altuve, Jose") %>% summarise(avg_abs_bot = mean(abs_sz_bot))

# graph density of pitch locations for Judge and Altuve with overlaid abs strike zone and regular strike zone

judge_pitch_loc_plot <- ggplot(full_pitch_df %>% filter(batter_name=="Judge, Aaron"), aes(x = plate_x, y = plate_z)) +
  stat_density2d(
    aes(fill = after_stat(level)),
    geom = "polygon",
    alpha = 0.5
  ) +
  scale_fill_continuous(name = "Density") +  geom_rect(
    xmin = -8.5, xmax = 8.5,
    ymin = judge_avg_abs_bot$avg_abs_bot, ymax = judge_avg_abs_top$avg_abs_top,
    fill = NA, color = "red",
    inherit.aes = FALSE
  ) +
  coord_fixed() +
  labs(title = "Aaron Judge Pitch Location Density, ABS Strike Zone Overlay", x = "Horizontal location (inches)", y = "Vertical location (inches)")

altuve_pitch_loc_plot <- ggplot(full_pitch_df %>% filter(batter_name=="Altuve, Jose"), aes(x = plate_x, y = plate_z)) +
  stat_density2d(
    aes(fill = after_stat(level)),
    geom = "polygon",
    alpha = 0.5
  ) +
  scale_fill_continuous(name = "Density") +  geom_rect(
    xmin = -8.5, xmax = 8.5,
    ymin = altuve_avg_abs_bot$avg_abs_bot, ymax = altuve_avg_abs_top$avg_abs_top,
    fill = NA, color = "red",
    inherit.aes = FALSE
  ) +
  coord_fixed() +
  labs(title = "Jose Altuve Pitch Location Density, ABS Strike Zone Overlay", x = "Horizontal location (inches)", y = "Vertical location (inches)")

judge_pitch_loc_plot
altuve_pitch_loc_plot
```
It seems like Judge may have seen more pitches outside of the abs strike zone than Altuve (higher density ring sticking outside abs zone). What proportion of pitches in the old strike zone will now be outside of the new strike zone for each player?

```{r player of interest 3}
# Add column "old_sz_biz" and "abs_sz_biz" to full_pitch_df
full_pitch_df <- full_pitch_df %>% mutate(old_sz_biz = ifelse(plate_z > sz_bot & plate_z < sz_top & plate_x > -8.5 & plate_x < 8.5, 1, 0), abs_sz_biz = ifelse(plate_z > abs_sz_bot & plate_z < abs_sz_top & plate_x > -8.5 & plate_x < 8.5, 1, 0))

 # What happened with those pitches?
full_pitch_df %>% filter(batter_name=="Judge, Aaron" & old_sz_biz == 1 & abs_sz_biz == 0) %>% nrow()

full_pitch_df %>% filter(batter_name=="Altuve, Jose" & old_sz_biz == 1 & abs_sz_biz == 0) %>% nrow()
```
So 118 pitches that were in Judge's old strike zone are now outside of the new strike zone, while only 74 pitches that were in Altuve's old strike zone are now outside of the new strike zone. What happened on those pitches?

```{r player of interest 4}
full_pitch_df %>% filter(batter_name=="Judge, Aaron" & old_sz_biz == 1 & abs_sz_biz == 0) %>% group_by(description)

full_pitch_df %>% filter(batter_name=="Altuve, Jose" & old_sz_biz == 1 & abs_sz_biz == 0) %>% nrow()
```


## Research Objectives:

#### Calculating strikezone differences for 2024:

```         
Start

  Mutate and create % strikezone diff
  
    Calculate average height x width for recorded strikezone and compare % area difference to ABS strikezone.
    
  Mutate and create batting average on balls in zone (BABIZ) for old and new zones
  
    Filter for pitches located in each zone, respectively, and calculate average
  
  Repeat for interesting stats
  
    wOBA/BABIPIZ/...
    
  Compare players who benefit vs suffer
  
    Group_by, sort
    
End
```


#### Long-term: model player performances for 2024 if strikezone was ABS size:

```         
Start

  Create distance from  strikezone metric
  
    ball_loc-nearest strikezone edge
    
  Train model on distance from strikezone edge, count, more etc
  
    Start with metrics like BA, can model other stats as well, ~ distance_from_zone, count, pitch velo, pitch break, etc
    
  Mutate predicted field to be NA for each pitch that is on the line between old and new zone
  
    filter for pitch location in ABS-Zone distance away from zone, make NA
    
  Predict outcome for those limbo-state pitches with new strike zone
  
  Make sure model is trained so if a batter would swing at a pitch on the edge of old zone, their habits stay true (and same if not swinging)
  
  Calculate difference in WAR, other key metrics for players. Who performed better? Worse?
  
End
```
