---
title: "ABS Strike Zone Comparison EDA"
author: "Charlie Braverman, Beili Chou, Andrew Hack"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Major League Baseball continues embracing technology by implementing an automated ball-strike (ABS) challenge system in spring training this year. While it will not be used in 2025, this system could be implemented in real games as soon as next year.. With a robotic strike zone, the league has changed its definition of the strike zone’s height to be 27% to 53.5% of a batter’s height. This marks a departure from the old rule from the midpoint between the shoulders and the top of the uniform pants to the top of the knees. While the 2025 season’s strike zone may be less impacted because human umpires will still be responsible for calling balls and strikes, the new strike zone may impact different players to different degrees. A key question for players, managers, and MLB front offices arises: which players stand to benefit from this change, and who might suffer?

The most obvious factor to investigate first is height. A taller player could see their effective strike zone shift more dramatically than a shorter counterpart. However, other elements, such as batting stance or body proportions, may also influence hitter and pitcher outcomes. By analyzing historical pitch-tracking data from Statcast, we will try to predict how the new strike zone rules will affect both hitters and pitchers.

## Conventions
**Variables** use `snake_case`.
**Functions** are verbs, e.g. `calculate_metrics()`.
**ABS** = Automated Ball-Strike challenge system.
**MLB** = Major League Baseball.
**wOBA** = Weighted On-Base Average.

## Packages & Functions

Uncommon Packages: baseballr @baseballr

baseballr is a package written for R focused on baseball analysis. It
includes functions for scraping various data from websites, such as
FanGraphs.com, Baseball-Reference.com, and baseballsavant.mlb.com. It
also includes functions for calculating metrics, such as wOBA, FIP, and
team-level consistency over custom time frames.

You can read more about some of the functions and how to use them at its
official site, <http://billpetti.github.io/baseballr/index.html>.

Sourced Functions: scrape_statcast() https://jacobrichey.github.io/2020-06-06-Build-a-Statcast-Database-in-R/

The scrape_statcast() function is a custom function that scrapes pitch-by-pitch data from Statcast for a given season. It takes a season as an argument and returns a data frame with all the pitch-by-pitch data for that season. The function uses the baseballr package to scrape the data from baseballsavant.mlb.com. This is necessary since baseballr scraping functions only return up to 25000 rows of data, and we need to scrape all the data for a season.

```{r load pkgs, warning=FALSE, message=FALSE}
library(tidyverse)
library(baseballr)
```


```{r load functions, warning=FALSE, message=FALSE}
# ──────────────────────────────────────────────────────────────────────────────
# Function: scrape_statcast()
# Purpose:  Scrapes full-season pitch-by-pitch data from Statcast, week by week.
#           Uses safely() to skip failed weeks and binds successful results.
# Args:
#   season: numeric or string (e.g., 2024)
# Returns:
#   A tibble of combined pitch data for the season.
# ──────────────────────────────────────────────────────────────────────────────
scrape_statcast <- function(season) {
  
  # 1) Build weekly date grid (Mar 25 → Nov 15)
  dates <- seq.Date(
    as.Date(paste0(season, '-03-25')),
    as.Date(paste0(season, '-11-15')),
    by = 'week'
  )
  date_grid <- tibble(start_date = dates, end_date = dates + 6)
  
  # 2) Wrap the savant scraper in safely() to catch errors
  safe_savant <- safely(scrape_statcast_savant_pitcher_all)
  
  # 3) Loop over each week, scraping via safe_savant()
  payload <- map(
    seq_along(date_grid$start_date), 
    ~{
      message('\nScraping week of ', date_grid$start_date[.x], '...\n')
      safe_savant(
        start_date = date_grid$start_date[.x], 
        end_date   = date_grid$end_date[.x]
      )
    }
  )
  
  # 4) Extract only successful results
  payload_df <- map(payload, 'result')
  nonempty <- map_df(
    seq_along(payload_df),
    ~tibble(week = .x, rows = nrow(payload_df[[.x]]))
  ) %>% 
    filter(rows > 0) %>% 
    pull(week)
  cleaned <- payload_df[nonempty]
  
  # 5) Bind all weekly tibbles into one
  bind_rows(cleaned)
}
```
## Scraping the data from Statcast, combining with player ID data, and biofile data.

Statcast provides pitch-by-pitch data for every game played in MLB. Our
approach will be to pull pitch data and join with another data set that
holds: batter_id, batter_name, and batter_height. This full data set
will allow for us to sort by batter produce the new rulebook strike zone
for each. 

Further documentation on variable names and their meanings can be found here: https://baseballsavant.mlb.com/csv-docs

Biofile data is obtained from https://www.retrosheet.org/biofile.htm, specifically the biofile0 since it contains heights in inches.

The output data frame will include all 113 variables from the Statcast data, as well as the batter's height in inches, strike zone top/bottom converted to inches, and the new ABS zone top/bottom in inches for each batter based on their height (Only the first 100,000 rows will be used in the CSV in GitHub due to size limitations, full_pitch_df object will be complete in environment).

One assumption that must be made is that batting stance will not affect the strike zone. This is a simplification that will be made for the sake of this analysis. In the future, maybe Statcast data will be used to record batting stance height, and this assumption can be revisited.

```{r load data, results='hide', collapse=TRUE, echo=TRUE, warning = FALSE, message = FALSE}
# Full Data Load & Join
# 1) Scrape 2024 Statcast data (may take ~25–30 minutes)
# 2) Load player ID lookup and join to get batter names
# 3) Load biofile heights and merge to get batter_height
# 4) Filter out missing zones and convert to inches
# 5) Compute ABS zone bounds, zone heights, areas, and in-zone flags
# 6) Create batting_team flag
# 7) Save a 50k-row sample for faster iteration
# ──────────────────────────────────────────────────────────────────────────────

t <- Sys.time()

# 1) Scrape Statcast
full_pitch_df <- scrape_statcast(season = 2024)

# 2) Player ID lookup
player_id_key_df <- chadwick_player_lu() %>%
  select(key_mlbam, key_retro, name_last, name_first) %>%
  filter(!is.na(key_mlbam))

full_pitch_df <- full_pitch_df %>%
  left_join(player_id_key_df, by = c("batter" = "key_mlbam")) %>%
  mutate(batter_name = paste(name_last, name_first, sep = ", "))

# 3) Batter heights from biofile
biofile_df <- read_csv("data/biofile_df.csv") %>% select(id, height)
batter_height_df <- full_pitch_df %>%
  full_join(biofile_df, by = c("key_retro" = "id")) %>%
  select(batter, height) %>% distinct()

full_pitch_df <- full_pitch_df %>%
  left_join(batter_height_df, by = "batter") %>%
  rename(batter_height = height)

# 4) Filter & convert to inches
full_pitch_df <- full_pitch_df %>%
  filter(!is.na(sz_top), !is.na(sz_bot)) %>%
  mutate(
    sz_bot  = sz_bot  * 12,
    sz_top  = sz_top  * 12,
    plate_x = plate_x * 12,
    plate_z = plate_z * 12
  )

# 5) Compute ABS zone & derived metrics
full_pitch_df <- full_pitch_df %>%
  mutate(
    abs_sz_bot    = 0.27 * batter_height,
    abs_sz_top    = 0.535 * batter_height,
    sz_height     = sz_top - sz_bot,
    abs_sz_height = abs_sz_top - abs_sz_bot,
    sz_area       = sz_height * 17,
    abs_sz_area   = abs_sz_height * 17,
    old_sz_biz    = ifelse(plate_z > sz_bot  & plate_z < sz_top &
                           plate_x > -8.5   & plate_x < 8.5, 1, 0),
    abs_sz_biz    = ifelse(plate_z > abs_sz_bot & plate_z < abs_sz_top &
                           plate_x > -8.5    & plate_x < 8.5, 1, 0)
  )

# 6) Batting team flag
full_pitch_df <- full_pitch_df %>%
  mutate(batting_team = ifelse(inning_topbot == "Top", away_team, home_team))

# 7) Save sample for iteration
sample_df_for_csv <- full_pitch_df[100000:150000, ]
write.csv(sample_df_for_csv, "data/full_pitch_df.csv")

# Calculate average zone bounds for plotting
avg_top <- mean(full_pitch_df$sz_top)
avg_bot <- mean(full_pitch_df$sz_bot)

Sys.time() - t
```

### Example outputs with the full data set.

Taking a look at pitch location data, which is broken down into x (horizontal) and z (vertical) coordinates from the catcher's POV, with a 2d density heat map. A 2D density heat map aggregates tens of thousands of pitch coordinates into smooth contours that reveal where pitchers most frequently target without suffering from over-plotting. By layering strike-zone boundaries on top, you can immediately compare pitch concentration inside and outside the zone. The color gradient encodes relative volume, making it easy to summarize location and frequency and to compare different zones or pitch types in a single intuitive graphic:

```{r pitch loc descriptives, warning=FALSE}
# ──────────────────────────────────────────────────────────────────────────────
# Pitch Location Density
#   • 2D density heatmap of plate_x vs plate_z
#   • Overlay of average umpire zone (red rect)
# ──────────────────────────────────────────────────────────────────────────────
pitch_loc_plot <- ggplot(full_pitch_df, aes(x = plate_x, y = plate_z)) +
  stat_density2d(aes(fill = after_stat(level)), geom = "polygon", alpha = 0.5) +
  scale_fill_continuous(name = "Density") +
  geom_rect(xmin = -8.5, xmax = 8.5, ymin = avg_bot, ymax = avg_top,
            fill = NA, color = "red", inherit.aes = FALSE) +
  coord_fixed() +
  labs(
    title = "Pitch Location Density",
    x     = "Horizontal location (inches)",
    y     = "Vertical location (inches)"
  )

pitch_loc_plot

summary(full_pitch_df$plate_x)
summary(full_pitch_df$plate_z)
```

And taking a look at how much the strike zone height varies across different batters:

```{r strike zone descriptive, warning=FALSE}
# ──────────────────────────────────────────────────────────────────────────────
# Strike Zone Height vs Batter Height
#   • Scatter + GAM smooth of current zone height
#   • No ABS overlay here; just descriptive.
# ──────────────────────────────────────────────────────────────────────────────
sz_height_plot <- ggplot(full_pitch_df, aes(x = batter_height, y = sz_height)) +
  geom_point() +
  geom_smooth(method = "gam") +
  labs(
    title = "Height vs. Strikezone Height",
    x     = "Batter height (inches)",
    y     = "Strikezone height (inches)"
  )

sz_height_plot
summary(full_pitch_df$sz_top)
summary(full_pitch_df$sz_bot)
```

## Exploring the data with ABS in mind.

Some initial thoughts:

How does the new strike zone compare to the old?

#### Exploring strikezone sizes:

```{r new strikezone, warning=FALSE}
# ──────────────────────────────────────────────────────────────────────────────
# Grouped summary & plot:
# 1) Compute average old vs ABS zone bounds by batter height
# 2) Plot old zone height vs batter_height, dashed line = ABS height
# ──────────────────────────────────────────────────────────────────────────────
sz_height_df <- full_pitch_df %>%
  group_by(batter_height) %>%
  summarise(
    avg_abs_sz_top    = mean(abs_sz_top),
    avg_abs_sz_bot    = mean(abs_sz_bot),
    avg_sz_top        = mean(sz_top),
    avg_sz_bot        = mean(sz_bot),
    avg_sz_height_diff= mean(sz_height) - mean(abs_sz_height)
  )

knitr::kable(sz_height_df)

sz_height_vs_height <- ggplot(full_pitch_df, aes(x = batter_height, y = sz_height)) +
  geom_point() +
  geom_smooth(method = "gam") +
  geom_line(aes(x = batter_height, y = abs_sz_height),
            linetype = "dashed") +
  labs(
    title = "Height vs. Strikezone Height",
    x     = "Batter height (inches)",
    y     = "Strikezone height (inches)"
  )

sz_height_vs_height
```

The new ABS zone is proportional to batter height, so it acts linearly. One takeaway is that batting stance can explain the deviation in strike zone height for each batter. However, an interesting note is that the ABS strike zone height is significantly smaller than the average current strike zone height at each batter height. Batting stance should theoretically cause current strike zone height to be smaller than ABS strike zone height, but this is not the case. Is the codification of the strike zone going to be a massive change for batters?

Not all batters and pitchers will be affected equally by the new strike zone. Some batters may benefit from the new strike zone, while others may suffer. We can calculate the difference in strike zone area between the current and ABS strike zones to see how much each batter's strike zone will change. We can also calculate the batting average on balls in zone (BABIZ) for each batter in the current and ABS strike zones to see how their performance might change. Obviously, players may change their approach to hitting based on the new strike zone, but this will give us a starting point to see who might benefit or suffer the most. 

#### Looking at "Players of Interest" Judge & Altuve.

Some players are notorious for their height, such as Aaron Judge and Jose Altuve. We can look at how their strike zones will change and how their performance might be affected.

```{r player of interest, warning=FALSE}
# ──────────────────────────────────────────────────────────────────────────────
# Compute and plot density + zone overlays for Aaron Judge & José Altuve
# ──────────────────────────────────────────────────────────────────────────────

judge_avg_abs_top <- full_pitch_df %>% filter(batter_name=="Judge, Aaron") %>% summarise(avg_abs_top = mean(abs_sz_top))
judge_avg_abs_bot <- full_pitch_df %>% filter(batter_name=="Judge, Aaron") %>% summarise(avg_abs_bot = mean(abs_sz_bot))
judge_avg_sz_top <- full_pitch_df %>% filter(batter_name=="Judge, Aaron", sz_top!="NA") %>% summarise(avg_sz_top = mean(sz_top))
judge_avg_sz_bot <- full_pitch_df %>% filter(batter_name=="Judge, Aaron", sz_bot!="NA") %>% summarise(avg_sz_bot = mean(sz_bot))

altuve_avg_abs_top <- full_pitch_df %>% filter(batter_name=="Altuve, José") %>% summarise(avg_abs_top = mean(abs_sz_top))
altuve_avg_abs_bot <- full_pitch_df %>% filter(batter_name=="Altuve, José") %>% summarise(avg_abs_bot = mean(abs_sz_bot))
altuve_avg_sz_top <- full_pitch_df %>% filter(batter_name=="Altuve, José") %>% summarise(avg_sz_top = mean(sz_top))
altuve_avg_sz_bot <- full_pitch_df %>% filter(batter_name=="Altuve, José") %>% summarise(avg_sz_bot = mean(sz_bot))

judge_pitch_loc_plot <- ggplot(full_pitch_df %>% filter(batter_name=="Judge, Aaron"), aes(x = plate_x, y = plate_z)) +
  stat_density2d(
    aes(fill = after_stat(level)),
    geom = "polygon",
    alpha = 0.5
  ) +
  scale_fill_continuous(name = "Density") +  geom_rect(
    xmin = -8.5, xmax = 8.5,
    ymin = judge_avg_abs_bot$avg_abs_bot, ymax = judge_avg_abs_top$avg_abs_top,
    fill = NA, color = "green",
    inherit.aes = FALSE
  ) + 
  geom_rect(
    xmin = -8.5, xmax = 8.5,
    ymin = judge_avg_sz_bot$avg_sz_bot, ymax = judge_avg_sz_top$avg_sz_top,
    fill = NA, color = "red",
    inherit.aes = FALSE
  ) +
  coord_fixed() +
  labs(title = "Aaron Judge Pitch Location Density, \nABS(green)/Old(red) Strike Zone Overlay", x = "Horizontal location (inches)", y = "Vertical location (inches)")

altuve_pitch_loc_plot <- ggplot(full_pitch_df %>% filter(batter_name=="Altuve, José"), aes(x = plate_x, y = plate_z)) +
  stat_density2d(
    aes(fill = after_stat(level)),
    geom = "polygon",
    alpha = 0.5
  ) +
  scale_fill_continuous(name = "Density") +  geom_rect(
    xmin = -8.5, xmax = 8.5,
    ymin = altuve_avg_abs_bot$avg_abs_bot, ymax = altuve_avg_abs_top$avg_abs_top,
    fill = NA, color = "green",
    inherit.aes = FALSE
  ) +
  geom_rect(
    xmin = -8.5, xmax = 8.5,
    ymin = altuve_avg_sz_bot$avg_sz_bot, ymax = altuve_avg_sz_top$avg_sz_top,
    fill = NA, color = "red",
    inherit.aes = FALSE
  ) +
  coord_fixed() + 
  labs(title = "José Altuve Pitch Location Density, \nABS(green)/Old(red) Strike Zone Overlay", x = "Horizontal location (inches)", y = "Vertical location (inches)")

judge_pitch_loc_plot
altuve_pitch_loc_plot
```

What number of pitches in the old strike zone will now be outside of the new strike zone for each player?

```{r player of interest 3}
# ──────────────────────────────────────────────────────────────────────────────
# Count pitches excluded by ABS zone for each player
# ──────────────────────────────────────────────────────────────────────────────
full_pitch_df %>%
  filter(batter_name=="Judge, Aaron", old_sz_biz == 1, abs_sz_biz == 0) %>%
  nrow()

full_pitch_df %>%
  filter(batter_name=="Altuve, José", old_sz_biz == 1, abs_sz_biz == 0) %>%
  nrow()
```

So (on a sample of data as an example) 121 pitches that were in Judge's old strike zone are now outside of the new strike zone, while only 73 pitches that were in Altuve's old strike zone are now outside of the new strike zone. What happened on those pitches?

```{r player of interest 4}
# ──────────────────────────────────────────────────────────────────────────────
# Tabulate outcomes on those excluded pitches
# ──────────────────────────────────────────────────────────────────────────────
full_pitch_df %>% filter(batter_name=="Judge, Aaron" & old_sz_biz == 1 & abs_sz_biz == 0) %>% group_by(description) %>% summarise(Outcome = n()) %>% arrange(desc(Outcome))

full_pitch_df %>% filter(batter_name=="Judge, Aaron" & old_sz_biz == 1 & abs_sz_biz == 0 & description == "hit_into_play") %>% group_by(events) %>% summarise(Outcome = n()) %>% arrange(desc(Outcome))


full_pitch_df %>% filter(batter_name=="Altuve, José" & old_sz_biz == 1 & abs_sz_biz == 0) %>% group_by(description) %>% summarise(Outcome = n()) %>% arrange(desc(Outcome))

full_pitch_df %>% filter(batter_name=="Altuve, José" & old_sz_biz == 1 & abs_sz_biz == 0 & description == "hit_into_play") %>% group_by(events) %>% summarise(Outcome = n()) %>% arrange(desc(Outcome))
```
So, for Judge, the most common outcome of pitches that were in his old strike zone but are now outside of the new strike zone is a foul ball, followed by swinging strikes. For Altuve, the most common outcome is a batted ball. On pitches put in play that meet the above criteria, Judge was a worse hitter than Altuve (.214AVG vs .296AVG on batted balls, respectively). With the change in strike zone, Judge would have to swing less at pitches that are now outside of the new strike zone, while Altuve *could* swing at those pitches if he hits them well, but then the change would be less impactful for him. 

Which teams had the best and worst [wOBA](https://www.lookoutlanding.com/2017/3/7/14783982/an-idiots-guide-to-advanced-statistics-woba-and-wrc-sabermetrics) on pitches in the old strike zone that would not be in the new strike zone?

```{r team of interest}
# ──────────────────────────────────────────────────────────────────────────────
# Team-level wOBA on pitches excluded by ABS zone
# ──────────────────────────────────────────────────────────────────────────────
team_wOBA_df <- full_pitch_df %>%
  filter(
    old_sz_biz == 1,
    abs_sz_biz == 0,
    description == "hit_into_play",
    estimated_woba_using_speedangle != "NA"
  ) %>%
  group_by(batting_team) %>%
  summarise(team_est_wOBA = mean(estimated_woba_using_speedangle), n = n()) %>%
  arrange(desc(team_est_wOBA))

knitr::kable(team_wOBA_df)
```

Interpretation of above could be: teams that have the highest wOBA on pitches that are now outside of the new strike zone (COL, MIL, PIT, ATL, PHI) are going to see less benefit from the new strike zone. Teams that have the lowest wOBA on pitches that are now outside of the new strike zone (LAD, CHC, CWS, TB, MIN) are going to see more benefit from the new strike zone.


## Initial research objectives to act as a guide to further research:

### Who will benefit the most? Who will suffer the most?

One question raised during this period of EDA was "how will the change in strike zone size affect players?" This broad question lends itself to a myriad of further investigation. To focus on identifying important questions to research, we brainstormed who the stakeholders in such a project would be. Players who could benefit from the change in strike zone size would be the most interested in the results of this analysis. Players who could suffer from the change in strike zone size would also be interested in the results. Managers and front offices would be interested in the results to determine how to adjust their strategies for the upcoming season. Fans would be interested in the results to see how their favorite players might be affected by the change in strike zone size.

The more fine grained questions about specific players' mechanics is too narrow a scope for the general body of stakeholders. However, the question of "who will benefit the most?" and "who will suffer the most?" is a good starting point for further research. 

The next steps will be to determine which metrics and interpretations of our EDA will be most useful for answering these questions. This will be done through a "research conversation" with the Sabermetrics community. Reading similar analyses, such as those done by Fangraphs, Baseball Prospectus, and other baseball analytics websites, will be helpful in determining the best approach to answering these questions.

## Exploring possible ideas for further research with Pseudocode

### Model player performances for 2024 if strikezone was ABS size:
To extend our research into a predictive framework, we could build a model that simulates how each batter might have performed in 2024 if the ABS strike zone had been in effect. The idea is to use pitch-by-pitch data to learn the relationship between a pitch’s proximity to the zone edge and various outcomes (e.g., batting average, strike rate), then “rerun” all the pitches that sit in the grey area between the old and new zones through that model to estimate new results. Below is a high-level pseudocode sketch of this process:

```         
Start

  Create distance fromstrikezone metric
  
    ball_loc-nearest strikezone edge
    
  Train model on distance from strikezone edge, count, more etc
  
    Start with metrics like BA, can model other stats as well, ~ distance_from_zone, count, pitch velo, pitch break, etc
    
  Mutate predicted field to be NA for each pitch that is on the line between old and new zone
  
    filter for pitch location in ABS-Zone distance away from zone, make NA
    
  Predict outcome for those limbo-state pitches with new strike zone
  
  Make sure model is trained so if a batter would swing at a pitch on the edge of old zone, their habits stay true (and same if not swinging)
  
  Calculate difference in WAR, other key metrics for players. Who performed better? Worse?
  
End
```
