---
title: "Presentation_Visuals"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

This notebook, **Presentation_Visuals**, is designed to produce the key figures and tables for our upcoming research presentation on how the proposed Automated Ball-Strike (ABS) strike zone would have altered MLB hitters’ outcomes in 2024.  

We begin by loading and preparing the pitch-by-pitch Statcast data, enriched with batter heights, and then:

1. **Visualize pitch distributions** under the current and ABS zones.  
2. **Compare aggregate hitting metrics** (e.g. wOBA, run expectancy, launch speed) between the old and new strike zones.  
3. **Isolate “limbo” pitches**—those in the old zone but excluded by ABS—and predict their impact on performance.  
4. **Highlight individual players** (Aaron Judge vs. José Altuve) to illustrate extremes of height-driven effects.  
5. **Rank top and bottom performers** (both players and teams) by the change in key metrics under ABS.  

Each section is self-contained: brief narrative, the data filtering or transformation step, and a ready-to-use ggplot or table.  The output is optimized for slide export, with clear titles and annotations.  

## Conventions
**Variables** use `snake_case`.
**Functions** are verbs, e.g. `calculate_metrics()`.
**ABS** = Automated Ball-Strike challenge system.
**MLB** = Major League Baseball.
**wOBA** = Weighted On-Base Average.
## Packages & Functions

Uncommon Packages: baseballr @baseballr

baseballr is a package written for R focused on baseball analysis. It
includes functions for scraping various data from websites, such as
FanGraphs.com, Baseball-Reference.com, and baseballsavant.mlb.com. It
also includes functions for calculating metrics, such as wOBA, FIP, and
team-level consistency over custom time frames.

You can read more about some of the functions and how to use them at its
official site, <http://billpetti.github.io/baseballr/index.html>.

Sourced Functions: scrape_statcast() https://jacobrichey.github.io/2020-06-06-Build-a-Statcast-Database-in-R/

The scrape_statcast() function is a custom function that scrapes pitch-by-pitch data from Statcast for a given season. It takes a season as an argument and returns a data frame with all the pitch-by-pitch data for that season. The function uses the baseballr package to scrape the data from baseballsavant.mlb.com. This is necessary since baseballr scraping functions only return up to 25000 rows of data, and we need to scrape all the data for a season.

```{r load pkgs, warning=FALSE, message=FALSE}
library(tidyverse)
library(baseballr)
library(kableExtra)
```


```{r load functions, warning=FALSE, message=FALSE}
# scrape data from Statcast and write to csv
scrape_statcast <- function(season) {
  
  # create weeks of dates for season from mar - nov
  # includes spring training + postseason if running from 01-01 to 12-31
  dates <- seq.Date(as.Date(paste0(season, '-03-25')),
                    as.Date(paste0(season, '-11-15')), by = 'week')
  
  date_grid <- tibble(start_date = dates, 
                      end_date = dates + 6)
  
  # create 'safe' version of scrape_statcast_savant in case week doesn't process
  safe_savant <- safely(scrape_statcast_savant_pitcher_all)
  
  # loop over each row of date_grid, and collect each week in a df
  payload <- map(.x = seq_along(date_grid$start_date), 
                 ~{message(paste0('\nScraping week of ', date_grid$start_date[.x], '...\n'))
                   
                   payload <- safe_savant(start_date = date_grid$start_date[.x], 
                                          end_date = date_grid$end_date[.x])
                   
                   return(payload)
                 })
  
  payload_df <- map(payload, 'result')
  
  # eliminate results with an empty dataframe
  number_rows <- map_df(.x = seq_along(payload_df), 
                        ~{number_rows <- tibble(week = .x, 
                                                number_rows = length(payload_df[[.x]]$game_date))}) %>%
    filter(number_rows > 0) %>%
    pull(week)
  
  payload_df_reduced <- payload_df[number_rows]
  
  combined <- payload_df_reduced %>%
    bind_rows()
  
  return(combined)
}
```
## Data

Statcast provides pitch-by-pitch data for every game played in MLB. Our
approach will be to pull pitch data and join with another data set that
holds: batter_id, batter_name, and batter_height. This full data set
will allow for us to sort by batter produce the new rulebook strike zone
for each. 

Further documentation on variable names and their meanings can be found here: https://baseballsavant.mlb.com/csv-docs

Biofile data is obtained from https://www.retrosheet.org/biofile.htm, specifically the biofile0 since it contains heights in inches.

The output data frame will include all 113 variables from the Statcast data, as well as the batter's height in inches, strike zone top/bottom converted to inches, and the new ABS zone top/bottom in inches for each batter based on their height (Only the first 100,000 rows will be used in the CSV in GitHub due to size limitations, full_pitch_df object will be complete in environment).

One assumption that must be made is that batting stance will not affect the strike zone. This is a simplification that will be made for the sake of this analysis. In the future, maybe Statcast data will be used to record batting stance height, and this assumption can be revisited.

```{r load data, results='hide', collapse=TRUE, echo=TRUE, warning = FALSE, message = FALSE}
# ONLY RUN THIS IF SCRAPING A FULL SEASON AND NOT JUST USING SAMPLE DATA
# tracking time for optimizing code efficiency
# If scraping a season, this chunk should take about 25-30 minutes to run (oof).
t <- Sys.time()

# scrape data from Statcast for 2024 season (Only run if data is not present in data file, or if you need to download another season.)
full_pitch_df <- scrape_statcast(season = 2024)


player_id_key_df <- chadwick_player_lu()

player_id_key_df <- player_id_key_df %>% select(key_mlbam, key_retro, name_last, name_first) %>% filter(!is.na(key_mlbam))

# Perform a join on "batter" in full_pitch_df and "key_mlbam" in player_id_key_df
full_pitch_df <- full_pitch_df %>%
  left_join(player_id_key_df, by = c("batter" = "key_mlbam")) %>%
  # Create the combined name field
  mutate(
    batter_name = paste(name_last, name_first, sep = ", "),
   key_retro   = key_retro
  )



# turn list of key_retro into character
full_pitch_df$key_retro <- as.character(full_pitch_df$key_retro)

# read in biofile data and join with full_pitch_df to get batter height
biofile_df <- read_csv("data/biofile_df.csv") %>% select(id, height)

batter_height_df <- full_pitch_df %>% full_join(biofile_df, by = c("key_retro" = "id")) %>% select(batter, height) %>% distinct()

# join full_pitch_df with batter_height_df to get batter height in pitch data
full_pitch_df <- full_pitch_df %>% left_join(batter_height_df, by = c("batter" = "batter")) %>% rename(batter_height = height)

# remove rows with NA sz_top and sz_bot
full_pitch_df <- full_pitch_df %>% filter(sz_top!="NA", sz_bot!="NA")

# changing sz_bot and top and plate_x/z to inches
full_pitch_df$sz_bot <- full_pitch_df$sz_bot * 12
full_pitch_df$sz_top <- full_pitch_df$sz_top * 12

full_pitch_df$plate_x <- full_pitch_df$plate_x * 12
full_pitch_df$plate_z <- full_pitch_df$plate_z * 12

# calculate new strike zone top and bottom based on batter height
full_pitch_df <- full_pitch_df %>% mutate(abs_sz_bot = .27 * batter_height, abs_sz_top = .535 * batter_height)

# calculate sz height and abs sz height
full_pitch_df <- full_pitch_df %>% mutate(sz_height = sz_top - sz_bot, abs_sz_height = abs_sz_top - abs_sz_bot)

# calculate strike zone area for current and ABS strike zones for all players
full_pitch_df <- full_pitch_df %>% mutate(sz_area = sz_height * 17, abs_sz_area = abs_sz_height * 17)

# Add column "old_sz_biz" and "abs_sz_biz" to full_pitch_df
full_pitch_df <- full_pitch_df %>% mutate(old_sz_biz = ifelse(plate_z > sz_bot & plate_z < sz_top & plate_x > -8.5 & plate_x < 8.5, 1, 0), abs_sz_biz = ifelse(plate_z > abs_sz_bot & plate_z < abs_sz_top & plate_x > -8.5 & plate_x < 8.5, 1, 0))

# create column "batting_team" that if inning_topbot == "Top", then away_team, else home_team
full_pitch_df <- full_pitch_df %>% mutate(batting_team = ifelse(inning_topbot == "Top", away_team, home_team))

# grab first 50,000 rows for sample data
sample_df_for_csv <- full_pitch_df[100000:150000,]
write.csv(sample_df_for_csv, "data/full_pitch_df.csv")

# using sample data for working with the data, full_pitch_df will be used for the final analysis.
#full_pitch_df <- read_csv("data/full_pitch_df.csv")

# Calculate average sz_top and sz_bot
avg_top <- mean(full_pitch_df$sz_top)
avg_bot <- mean(full_pitch_df$sz_bot)

# end time tracking
Sys.time()-t
```


# Visuals

## old vs new strike zone

```{r old vs new sz, echo=FALSE}
avg_old_top <- mean(df_2024$sz_top, na.rm = TRUE)
avg_old_bot <- mean(df_2024$sz_bot, na.rm = TRUE)
avg_new_top <- mean(df_2024$abs_sz_top, na.rm = TRUE)
avg_new_bot <- mean(df_2024$abs_sz_bot, na.rm = TRUE)

pitch_loc_plot <- ggplot(full_pitch_df, aes(x = plate_x, y = plate_z)) +
  stat_density2d(
    aes(fill = after_stat(level)),
    geom = "polygon",
    alpha = 0.5
  ) +
  scale_fill_continuous(name = "Density") +  geom_rect(
    xmin = -8.5, xmax = 8.5,
    ymin = avg_new_bot, ymax = avg_new_top,
    fill = NA, color = "green",
    inherit.aes = FALSE
  ) + 
  geom_rect(
    xmin = -8.5, xmax = 8.5,
    ymin = avg_old_bot, ymax = avg_old_top,
    fill = NA, color = "red",
    inherit.aes = FALSE
  ) +
  coord_fixed() +
  labs(title = "The ABS zone is significantly smaller than the old strike zone", subtitle = "ABS (green) vs Old (red)", x = "Horizontal location (inches)", y = "Vertical location (inches)")

pitch_loc_plot
```

## old vs new stats comparison

```{r old vs new stats comparison, echo=FALSE}
df <- full_pitch_df %>%
  filter(plate_x >= -8.5, plate_x <= 8.5, description == "hit_into_play")

# Define the old strike zone: plate_z between sz_bot and sz_top.
old_zone_df <- df %>%
  filter(plate_z >= sz_bot, plate_z <= sz_top)

# Define the new ABS strike zone: plate_z between abs_sz_bot and abs_sz_top.
new_zone_df <- df %>%
  filter(plate_z >= abs_sz_bot, plate_z <= abs_sz_top)

# Calculate the mean values for key statistics in each zone.
stats_old <- old_zone_df %>%
  summarise(
    mean_woba = mean(estimated_woba_using_speedangle, na.rm = TRUE),
    mean_delta_run_exp = mean(delta_run_exp, na.rm = TRUE),
    mean_launch_speed = mean(launch_speed, na.rm = TRUE)
  )

stats_new <- new_zone_df %>%
  summarise(
    mean_woba = mean(estimated_woba_using_speedangle, na.rm = TRUE),
    mean_delta_run_exp = mean(delta_run_exp, na.rm = TRUE),
    mean_launch_speed = mean(launch_speed, na.rm = TRUE)
  )

# Combine the statistics into one data frame.
stats_comparison <- data.frame(
  Stat = c("Estimated wOBA", "Delta Run Exp", "Launch Speed"),
  Old = c(stats_old$mean_woba, stats_old$mean_delta_run_exp, stats_old$mean_launch_speed),
  New = c(stats_new$mean_woba, stats_new$mean_delta_run_exp, stats_new$mean_launch_speed)
)

# Pivot the data to long format for ggplot2.
stats_long <- stats_comparison %>%
  pivot_longer(cols = c("Old", "New"), names_to = "Zone", values_to = "MeanValue")

# Compute t-test p-values for each statistic.
p_val_woba <- t.test(old_zone_df$estimated_woba_using_speedangle, 
                     new_zone_df$estimated_woba_using_speedangle)$p.value
p_val_dre <- t.test(old_zone_df$delta_run_exp, 
                    new_zone_df$delta_run_exp)$p.value
p_val_launch <- t.test(old_zone_df$launch_speed, 
                       new_zone_df$launch_speed)$p.value

# Create a data frame with p-values and assign significance stars.
pvals <- data.frame(
  Stat = c("Estimated wOBA", "Delta Run Exp", "Launch Speed"),
  p_value = c(p_val_woba, p_val_dre, p_val_launch)
)
pvals$sig_label <- ifelse(pvals$p_value < 0.001, "***",
                    ifelse(pvals$p_value < 0.01, "**",
                    ifelse(pvals$p_value < 0.05, "*", "ns")))

# Determine a y-position for the annotations by getting the max mean value per stat.
max_vals <- stats_long %>% group_by(Stat) %>% summarise(max_val = max(MeanValue))
pvals <- merge(pvals, max_vals, by = "Stat")
pvals <- pvals %>% mutate(y_position = max_val * 1.1)

# Create the faceted grouped bar chart with custom fill colors and p-value annotations.
ggplot(stats_long, aes(x = Zone, y = MeanValue, fill = Zone)) +
  geom_col() +
  facet_wrap(~Stat, scales = "free_y") +
  scale_fill_manual(values = c("Old" = "red", "New" = "green")) +
  geom_text(data = pvals, 
            aes(x = 1.5, y = y_position, 
                label = paste("p =", signif(p_value, 2), sig_label)),
            inherit.aes = FALSE, size = 3) +
  labs(title = "Between old and new zones, stat changes across MLB are not significant",
       x = "Zone",
       y = "Mean Value") +
  theme_minimal()
```

## old but not new vs new

```{r old but not new vs new, echo=FALSE}
# Group 1: Old-Only (pitches in old strike zone but NOT in new ABS zone)
old_only_df <- df %>%
  filter(plate_z >= sz_bot, plate_z <= sz_top) %>%
  filter(!(plate_z >= abs_sz_bot & plate_z <= abs_sz_top))

# Group 2: New ABS (pitches in the new ABS strike zone)
new_zone_df <- df %>%
  filter(plate_z >= abs_sz_bot, plate_z <= abs_sz_top)

# Calculate the mean key statistics for each group
stats_old_only <- old_only_df %>%
  summarise(
    mean_woba = mean(estimated_woba_using_speedangle, na.rm = TRUE),
    mean_delta_run_exp = mean(delta_run_exp, na.rm = TRUE),
    mean_launch_speed = mean(launch_speed, na.rm = TRUE)
  )

stats_new <- new_zone_df %>%
  summarise(
    mean_woba = mean(estimated_woba_using_speedangle, na.rm = TRUE),
    mean_delta_run_exp = mean(delta_run_exp, na.rm = TRUE),
    mean_launch_speed = mean(launch_speed, na.rm = TRUE)
  )

# Combine the statistics into one data frame for comparison
stats_comparison <- data.frame(
  Stat = c("Estimated wOBA", "Delta Run Exp", "Launch Speed"),
  Old_Only = c(stats_old_only$mean_woba, stats_old_only$mean_delta_run_exp, stats_old_only$mean_launch_speed),
  New = c(stats_new$mean_woba, stats_new$mean_delta_run_exp, stats_new$mean_launch_speed)
)

# Pivot to long format for ggplot2
stats_long <- stats_comparison %>%
  pivot_longer(cols = c("Old_Only", "New"), names_to = "Group", values_to = "MeanValue")

# Compute t-test p-values for each statistic between the two groups
p_val_woba <- t.test(old_only_df$estimated_woba_using_speedangle, new_zone_df$estimated_woba_using_speedangle)$p.value
p_val_dre <- t.test(old_only_df$delta_run_exp, new_zone_df$delta_run_exp)$p.value
p_val_launch <- t.test(old_only_df$launch_speed, new_zone_df$launch_speed)$p.value

pvals <- data.frame(
  Stat = c("Estimated wOBA", "Delta Run Exp", "Launch Speed"),
  p_value = c(p_val_woba, p_val_dre, p_val_launch)
)
pvals$sig_label <- ifelse(pvals$p_value < 0.001, "***",
                    ifelse(pvals$p_value < 0.01, "**",
                    ifelse(pvals$p_value < 0.05, "*", "ns")))

# Determine a y-position for annotations by finding the max MeanValue for each stat
max_vals <- stats_long %>% group_by(Stat) %>% summarise(max_val = max(MeanValue))
pvals <- merge(pvals, max_vals, by = "Stat")
pvals <- pvals %>% mutate(y_position = max_val * 1.1)

# Create the faceted grouped bar chart with custom colors and p-value annotations
ggplot(stats_long, aes(x = Group, y = MeanValue, fill = Group)) +
  geom_col() +
  facet_wrap(~Stat, scales = "free_y") +
  scale_fill_manual(values = c("Old_Only" = "red", "New" = "green")) +
  geom_text(data = pvals, 
            aes(x = 1.5, y = y_position, 
                label = paste("p =", signif(p_value, 2), sig_label)),
            inherit.aes = FALSE, size = 3) +
  labs(title = "Batters hit significantly better in the new zone than area of old zone being cut off",
       x = "Group",
       y = "Mean Value") +
  theme_minimal()
```

## judge vs altuve

```{r judge vs altuve, echo=FALSE}
# 1. Filter data for the two batters, horizontal plate limits, and hit_into_play events.
df <- full_pitch_df %>%
  filter(plate_x >= -8.5, plate_x <= 8.5,
         description == "hit_into_play",
         batter_name %in% c("Judge, Aaron", "Altuve, José"))

# 2. Create a zone classification variable:
#    - "New" if plate_z is between abs_sz_bot and abs_sz_top,
#    - "Old_Only" if plate_z is between sz_bot and sz_top but not in the new zone.
df <- df %>%
  mutate(zone_group = case_when(
    plate_z >= abs_sz_bot & plate_z <= abs_sz_top ~ "New",
    plate_z >= sz_bot & plate_z <= sz_top ~ "Old_Only",
    TRUE ~ NA_character_
  )) %>%
  filter(!is.na(zone_group)) %>%
  mutate(zone_group = factor(zone_group, levels = c("Old_Only", "New")),
         batter_name = factor(batter_name, levels = c("Judge, Aaron", "Altuve, José"))
         )

# 3. Aggregate stats by batter and zone_group.
agg_stats <- df %>%
  group_by(batter_name, zone_group) %>%
  summarise(
    mean_woba = mean(estimated_woba_using_speedangle, na.rm = TRUE),
    mean_dre = mean(delta_run_exp, na.rm = TRUE),
    mean_launch = mean(launch_speed, na.rm = TRUE)
  ) %>%
  ungroup()

# Pivot to wide format so we have one row per batter.
agg_stats_wide <- agg_stats %>%
  pivot_wider(names_from = zone_group, values_from = c(mean_woba, mean_dre, mean_launch))

# 4. Compute the difference (New - Old_Only) for each stat and each batter.
stats_diff <- agg_stats_wide %>%
  mutate(
    diff_woba = mean_woba_New - mean_woba_Old_Only,
    diff_dre = mean_dre_New - mean_dre_Old_Only,
    diff_launch = mean_launch_New - mean_launch_Old_Only
  ) %>%
  select(batter_name, diff_woba, diff_dre, diff_launch)

# Pivot to long format for plotting.
stats_long <- stats_diff %>%
  pivot_longer(cols = starts_with("diff_"), names_to = "Stat", values_to = "Difference") %>%
  mutate(Stat = case_when(
    Stat == "diff_woba" ~ "Estimated wOBA",
    Stat == "diff_dre" ~ "Delta Run Exp",
    Stat == "diff_launch" ~ "Launch Speed"
  ))

# 5. Create the faceted bar chart comparing the differences (New - Old_Only) for each batter,
#    adjusting the bar width and x-axis expansion to prevent overlapping labels.
ggplot(stats_long, aes(x = batter_name, y = Difference, fill = batter_name)) +
  geom_col(width = 0.6) +
  facet_wrap(~Stat, scales = "free_y") +
  scale_fill_manual(values = c("Judge, Aaron" = "navy", "Altuve, José" = "orange")) +
  scale_x_discrete(expand = expansion(mult = c(0.3, 0.3))) +
  labs(title = "Judge, a tall batter, benefits more from the new zone than Altuve, a shorter batter",
       x = "Batter",
       y = "Difference (New - Old)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## top 10 and bottom 10 players

```{r top10/bot10, echo=FALSE}
# 1. Filter the data: horizontal plate limits and "hit_into_play" events.
df <- full_pitch_df %>%
  filter(plate_x >= -8.5, plate_x <= 8.5,
         description == "hit_into_play")

# 2. Compute plate appearances per batter using unique game_pk and at_bat_number.
pa_counts <- df %>%
  group_by(batter_name) %>%
  summarise(pa = n_distinct(paste(game_pk, at_bat_number, sep = "_"))) 

# 3. Retain only batters with at least 300 plate appearances.
df <- df %>%
  inner_join(pa_counts, by = "batter_name") %>%
  filter(pa >= 300)

# 4. Classify each pitch as either "New" (if plate_z is between abs_sz_bot and abs_sz_top)
#    or "Old_Only" (if plate_z is between sz_bot and sz_top but not in the new zone).
df <- df %>%
  mutate(zone_group = case_when(
    plate_z >= abs_sz_bot & plate_z <= abs_sz_top ~ "New",
    plate_z >= sz_bot  & plate_z <= sz_top  ~ "Old_Only",
    TRUE ~ NA_character_
  )) %>%
  filter(!is.na(zone_group)) %>%
  mutate(zone_group = factor(zone_group, levels = c("Old_Only", "New")))

# 5. Aggregate key stats by batter (including batter_height) and zone_group.
agg_stats <- df %>%
  group_by(batter_name, batter_height, zone_group) %>%
  summarise(
    mean_woba   = mean(estimated_woba_using_speedangle, na.rm = TRUE),
    mean_dre    = mean(delta_run_exp, na.rm = TRUE),
    mean_launch = mean(launch_speed, na.rm = TRUE),
    .groups = "drop"
  )

# 6. Pivot to wide format so each batter has separate columns for Old_Only and New.
agg_stats_wide <- agg_stats %>%
  pivot_wider(names_from = zone_group, 
              values_from = c(mean_woba, mean_dre, mean_launch))

# 7. Compute the differences (New - Old_Only) for each metric.
agg_stats_wide <- agg_stats_wide %>%
  mutate(
    diff_woba   = mean_woba_New - mean_woba_Old_Only,
    diff_dre    = mean_dre_New - mean_dre_Old_Only,
    diff_launch = mean_launch_New - mean_launch_Old_Only
  )

# 8. Extract the top 10 and bottom 10 players for each metric.
# Delta Run Exp
top10_dre <- agg_stats_wide %>% 
  arrange(desc(diff_dre)) %>% 
  slice_head(n = 10)
bottom10_dre <- agg_stats_wide %>% 
  arrange(diff_dre) %>% 
  slice_head(n = 10)

# Estimated wOBA
top10_woba <- agg_stats_wide %>% 
  arrange(desc(diff_woba)) %>% 
  slice_head(n = 10)
bottom10_woba <- agg_stats_wide %>% 
  arrange(diff_woba) %>% 
  slice_head(n = 10)

# Launch Speed
top10_launch <- agg_stats_wide %>% 
  arrange(desc(diff_launch)) %>% 
  slice_head(n = 10)
bottom10_launch <- agg_stats_wide %>% 
  arrange(diff_launch) %>% 
  slice_head(n = 10)

# 9. For each subset, create a new column for batter label including height.
#    Also, compute the position for the stat label.
#    We'll use an offset equal to 5% of the maximum absolute difference in each subset.
add_labels <- function(df, diff_col) {
  df %>%
    mutate(batter_label = paste(batter_name, "(", batter_height, "in)", sep = " ")) %>%
    mutate(abs_diff = abs(!!rlang::sym(diff_col))) %>%
    mutate(max_abs = max(abs_diff, na.rm = TRUE)) %>%
    mutate(label_pos = ifelse(!!rlang::sym(diff_col) > 0,
                              !!rlang::sym(diff_col) - 0.05 * max_abs,
                              !!rlang::sym(diff_col) + 0.05 * max_abs),
           hjust = ifelse(!!rlang::sym(diff_col) > 0, 1, 0))
}

# Apply the helper function to each subset.
top10_dre <- add_labels(top10_dre, "diff_dre")
bottom10_dre <- add_labels(bottom10_dre, "diff_dre")
top10_woba <- add_labels(top10_woba, "diff_woba")
bottom10_woba <- add_labels(bottom10_woba, "diff_woba")
top10_launch <- add_labels(top10_launch, "diff_launch")
bottom10_launch <- add_labels(bottom10_launch, "diff_launch")

# 10. Create the six separate bar graphs.

# Graph 1: Top 10 Players by Delta Run Exp Difference
plot_top_dre <- ggplot(top10_dre, aes(x = reorder(batter_label, diff_dre), y = diff_dre)) +
  geom_col(fill = "forestgreen", width = 0.6) +
  geom_text(aes(label = sprintf("%.2f", diff_dre), y = label_pos, hjust = hjust), 
            color = "white", size = 3) +
  coord_flip() +
  labs(title = "Top 10 Players by Delta Run Exp Difference",
       x = "Batter (Height)", y = "Difference (New - Old)") +
  theme_minimal()

# Graph 2: Bottom 10 Players by Delta Run Exp Difference
plot_bottom_dre <- ggplot(bottom10_dre, aes(x = reorder(batter_label, diff_dre), y = diff_dre)) +
  geom_col(fill = "firebrick", width = 0.6) +
  geom_text(aes(label = sprintf("%.2f", diff_dre), y = label_pos, hjust = hjust), 
            color = "white", size = 3) +
  coord_flip() +
  labs(title = "Bottom 10 Players by Delta Run Exp Difference",
       x = "Batter (Height)", y = "Difference (New - Old)") +
  theme_minimal()

# Graph 3: Top 10 Players by Estimated wOBA Difference
plot_top_woba <- ggplot(top10_woba, aes(x = reorder(batter_label, diff_woba), y = diff_woba)) +
  geom_col(fill = "forestgreen", width = 0.6) +
  geom_text(aes(label = sprintf("%.2f", diff_woba), y = label_pos, hjust = hjust), 
            color = "white", size = 3) +
  coord_flip() +
  labs(title = "Top 10 Players by Estimated wOBA Difference",
       x = "Batter (Height)", y = "Difference (New - Old)") +
  theme_minimal()

# Graph 4: Bottom 10 Players by Estimated wOBA Difference
plot_bottom_woba <- ggplot(bottom10_woba, aes(x = reorder(batter_label, diff_woba), y = diff_woba)) +
  geom_col(fill = "firebrick", width = 0.6) +
  geom_text(aes(label = sprintf("%.2f", diff_woba), y = label_pos, hjust = hjust), 
            color = "white", size = 3) +
  coord_flip() +
  labs(title = "Bottom 10 Players by Estimated wOBA Difference",
       x = "Batter (Height)", y = "Difference (New - Old)") +
  theme_minimal()

# Graph 5: Top 10 Players by Launch Speed Difference
plot_top_launch <- ggplot(top10_launch, aes(x = reorder(batter_label, diff_launch), y = diff_launch)) +
  geom_col(fill = "forestgreen", width = 0.6) +
  geom_text(aes(label = sprintf("%.2f", diff_launch), y = label_pos, hjust = hjust), 
            color = "white", size = 3) +
  coord_flip() +
  labs(title = "Top 10 Players by Launch Speed Difference",
       x = "Batter (Height)", y = "Difference (New - Old)") +
  theme_minimal()

# Graph 6: Bottom 10 Players by Launch Speed Difference
plot_bottom_launch <- ggplot(bottom10_launch, aes(x = reorder(batter_label, diff_launch), y = diff_launch)) +
  geom_col(fill = "firebrick", width = 0.6) +
  geom_text(aes(label = sprintf("%.2f", diff_launch), y = label_pos, hjust = hjust), 
            color = "white", size = 3) +
  coord_flip() +
  labs(title = "Bottom 10 Players by Launch Speed Difference",
       x = "Batter (Height)", y = "Difference (New - Old)") +
  theme_minimal()

# Display the graphs
plot_top_dre
plot_bottom_dre
plot_top_woba
plot_bottom_woba
plot_top_launch
plot_bottom_launch
```

## top 10 and bottom 10 teams

```{r team charts, echo=FALSE}
# 1. Filter the data: only include pitches with plate_x between -8.5 and 8.5 and description "hit_into_play".
df_team <- full_pitch_df %>%
  filter(plate_x >= -8.5, plate_x <= 8.5,
         description == "hit_into_play")

# 2. Compute team-level plate appearances using unique combinations of game_pk and at_bat_number.
pa_counts_team <- df_team %>%
  group_by(batting_team) %>%
  summarise(pa = n_distinct(paste(game_pk, at_bat_number, sep = "_")))

# 3. Retain only teams with at least 300 plate appearances.
df_team <- df_team %>%
  inner_join(pa_counts_team, by = "batting_team") %>%
  filter(pa >= 300)

# 4. Classify each pitch as either "New" (if plate_z is between abs_sz_bot and abs_sz_top)
#    or "Old_Only" (if plate_z is between sz_bot and sz_top but not in the new zone).
df_team <- df_team %>%
  mutate(zone_group = case_when(
    plate_z >= abs_sz_bot & plate_z <= abs_sz_top ~ "New",
    plate_z >= sz_bot  & plate_z <= sz_top ~ "Old_Only",
    TRUE ~ NA_character_
  )) %>%
  filter(!is.na(zone_group)) %>%
  mutate(zone_group = factor(zone_group, levels = c("Old_Only", "New")))

# 5. Aggregate key stats by team and zone_group.
agg_stats_team <- df_team %>%
  group_by(batting_team, zone_group) %>%
  summarise(
    mean_woba   = mean(estimated_woba_using_speedangle, na.rm = TRUE),
    mean_dre    = mean(delta_run_exp, na.rm = TRUE),
    mean_launch = mean(launch_speed, na.rm = TRUE),
    .groups = "drop"
  )

# 6. Pivot the data to wide format so each team has separate columns for Old_Only and New.
agg_stats_team_wide <- agg_stats_team %>%
  pivot_wider(names_from = zone_group, 
              values_from = c(mean_woba, mean_dre, mean_launch))

# 7. Compute the differences (New - Old_Only) for each metric.
agg_stats_team_wide <- agg_stats_team_wide %>%
  mutate(
    diff_woba   = mean_woba_New - mean_woba_Old_Only,
    diff_dre    = mean_dre_New - mean_dre_Old_Only,
    diff_launch = mean_launch_New - mean_launch_Old_Only
  )

# 8. Create a label column (here, simply the team name).
agg_stats_team_wide <- agg_stats_team_wide %>%
  mutate(team_label = batting_team)

# 9. Extract the top 10 and bottom 10 teams for each metric.
top10_dre_team <- agg_stats_team_wide %>% arrange(desc(diff_dre)) %>% slice_head(n = 10)
bottom10_dre_team <- agg_stats_team_wide %>% arrange(diff_dre) %>% slice_head(n = 10)

top10_woba_team <- agg_stats_team_wide %>% arrange(desc(diff_woba)) %>% slice_head(n = 10)
bottom10_woba_team <- agg_stats_team_wide %>% arrange(diff_woba) %>% slice_head(n = 10)

top10_launch_team <- agg_stats_team_wide %>% arrange(desc(diff_launch)) %>% slice_head(n = 10)
bottom10_launch_team <- agg_stats_team_wide %>% arrange(diff_launch) %>% slice_head(n = 10)

# 10. Define a helper function to compute a label position for the numeric difference.
add_labels_team <- function(df, diff_col) {
  df %>%
    mutate(abs_diff = abs(!!rlang::sym(diff_col))) %>%
    mutate(max_abs = max(abs_diff, na.rm = TRUE)) %>%
    mutate(label_pos = ifelse(!!rlang::sym(diff_col) > 0,
                              !!rlang::sym(diff_col) - 0.05 * max_abs,
                              !!rlang::sym(diff_col) + 0.05 * max_abs),
           hjust = ifelse(!!rlang::sym(diff_col) > 0, 1, 0))
}

# Apply the helper function to each subset.
top10_dre_team <- add_labels_team(top10_dre_team, "diff_dre")
bottom10_dre_team <- add_labels_team(bottom10_dre_team, "diff_dre")

top10_woba_team <- add_labels_team(top10_woba_team, "diff_woba")
bottom10_woba_team <- add_labels_team(bottom10_woba_team, "diff_woba")

top10_launch_team <- add_labels_team(top10_launch_team, "diff_launch")
bottom10_launch_team <- add_labels_team(bottom10_launch_team, "diff_launch")

# 11. Create six separate graphs.

# Graph 1: Top 10 Teams by Delta Run Exp Difference
plot_top_dre_team <- ggplot(top10_dre_team, aes(x = reorder(team_label, diff_dre), y = diff_dre)) +
  geom_col(fill = "forestgreen", width = 0.6) +
  geom_text(aes(label = sprintf("%.2f", diff_dre), y = label_pos, hjust = hjust),
            color = "white", size = 3) +
  coord_flip() +
  labs(title = "Top 10 Teams by Delta Run Exp Difference",
       x = "Team", y = "Difference (New - Old)") +
  theme_minimal()

# Graph 2: Bottom 10 Teams by Delta Run Exp Difference
plot_bottom_dre_team <- ggplot(bottom10_dre_team, aes(x = reorder(team_label, diff_dre), y = diff_dre)) +
  geom_col(fill = "firebrick", width = 0.6) +
  geom_text(aes(label = sprintf("%.2f", diff_dre), y = label_pos, hjust = hjust),
            color = "white", size = 3) +
  coord_flip() +
  labs(title = "Bottom 10 Teams by Delta Run Exp Difference",
       x = "Team", y = "Difference (New - Old)") +
  theme_minimal()

# Graph 3: Top 10 Teams by Estimated wOBA Difference
plot_top_woba_team <- ggplot(top10_woba_team, aes(x = reorder(team_label, diff_woba), y = diff_woba)) +
  geom_col(fill = "forestgreen", width = 0.6) +
  geom_text(aes(label = sprintf("%.2f", diff_woba), y = label_pos, hjust = hjust),
            color = "white", size = 3) +
  coord_flip() +
  labs(title = "Top 10 Teams by Estimated wOBA Difference",
       x = "Team", y = "Difference (New - Old)") +
  theme_minimal()

# Graph 4: Bottom 10 Teams by Estimated wOBA Difference
plot_bottom_woba_team <- ggplot(bottom10_woba_team, aes(x = reorder(team_label, diff_woba), y = diff_woba)) +
  geom_col(fill = "firebrick", width = 0.6) +
  geom_text(aes(label = sprintf("%.2f", diff_woba), y = label_pos, hjust = hjust),
            color = "white", size = 3) +
  coord_flip() +
  labs(title = "Bottom 10 Teams by Estimated wOBA Difference",
       x = "Team", y = "Difference (New - Old)") +
  theme_minimal()

# Graph 5: Top 10 Teams by Launch Speed Difference
plot_top_launch_team <- ggplot(top10_launch_team, aes(x = reorder(team_label, diff_launch), y = diff_launch)) +
  geom_col(fill = "forestgreen", width = 0.6) +
  geom_text(aes(label = sprintf("%.2f", diff_launch), y = label_pos, hjust = hjust),
            color = "white", size = 3) +
  coord_flip() +
  labs(title = "Top 10 Teams by Launch Speed Difference",
       x = "Team", y = "Difference (New - Old)") +
  theme_minimal()

# Graph 6: Bottom 10 Teams by Launch Speed Difference
plot_bottom_launch_team <- ggplot(bottom10_launch_team, aes(x = reorder(team_label, diff_launch), y = diff_launch)) +
  geom_col(fill = "firebrick", width = 0.6) +
  geom_text(aes(label = sprintf("%.2f", diff_launch), y = label_pos, hjust = hjust),
            color = "white", size = 3) +
  coord_flip() +
  labs(title = "Bottom 10 Teams by Launch Speed Difference",
       x = "Team", y = "Difference (New - Old)") +
  theme_minimal()

# Display the graphs
plot_top_dre_team
plot_bottom_dre_team
plot_top_woba_team
plot_bottom_woba_team
plot_top_launch_team
plot_bottom_launch_team
```